---
title: "PairProgramming_Slicing"
author: "HDS"
date: "2024-08-09"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Pair Programming Slicing

Using dplyr to manipulate, structure and sort data frames

See Wickham et al,  Chapter 3

https://r4ds.hadley.nz/data-transform


HD Sheets, August 9, 2024
checked 01/03/2025

Student Name:Ryan Waterman
Teamate Name:Nicholas Perry

#Topics- by class of operations

These are tools to regroup data for various purposes, including plotting
and generating summary tables,  such as "cross tables" or "pivot tables" easily

Row Manipulations

  -filter()
  -arrange()
  -distinct()
  
Column Manipulations

  -mutate()
  -select()
  -rename()
  -relocate()

Grouping

  -group_by()

Summarize or calculate

  -summarize()

#Libraries

```{R}
library('tidyverse')
library('ggplot2')
```

#example data

we will use the rather standard mtcars dataset of car performance data
created by MotorTrend (MT) in the 1970s

Not super exciting, but small enough to work with

Read that help menu!

```{R}
data(mtcars)
head(mtcars)
help(mtcars)
```

#Examples of filtering

Selecting specific rows, based on some condition on a variable

Filter on one variable

```{R}
# 8 cylinder cars only

filter(mtcars,cyl==8)
```
#filter on a condition

using a pipeline %>% to feed the dataframe into the filter

looking for cars with less than 75 hp (yeah, really...)

```{R}
mtcars %>%filter(hp<75)
```

#combined conditions

Let's find 6 cylinder cars with mpg>25

we use an AND operation "&" to combine the conditions (cyl==6) with (mpg>20)

```{R}
mtcars %>%filter( (cyl==6)&(mpg>20))

```

Hmm, lousy fuel economy in the 70s for 6 cylinders.


#*Question/Action*

filter the cars with 6 cylinders and a manual transmission
the column am is 0 for automatic transmissions and 1 for manuals

```{R}
mtcars %>% filter((cyl==6)&(am==1))
```

# arrange()

This is a sorting function

sort by increasing mpg

```{R}
mtcars %>% arrange(mpg)
```
decreasing mpgh

Put desc() around the variable

```{R}
mtcars %>% arrange(desc(mpg))
```

##sorting on multiple variables

just include them all in the arrange(), using 
desc() as needed

Lets sort by cylinder (descending) and qsec time (increasing, low qsec time
means a fast car)

This looks a lot like the sorting options in Excel

```{R}
mtcars %>% arrange(desc(cyl),qsec)
```
#*Queston/Action*

sort the cars by transmission type, mpg and number of cylinders

```{r}
mtcars %>% arrange(am,mpg,cyl)
```

# distinct,  finding unique instances

of cylinders

This allows use to determine how many distinct or unique values we have in a column

```{R}
mtcars %>% distinct(cyl)
```

Let's look at combinations of cylinders and carbeurators

note the alternate form of the pipe |> is easier to type

```{R}
mtcars |> distinct(cyl,carb)
```

Okay, but I want it ordered by cyl

we just pipeline the arrange() after the distinct

```{R}
mtcars |> distinct(cyl,carb) |> arrange(cyl,carb)
```

That is much easier to read

#*Question/Action*

find the unique combinations of gear and carb and am appearing in the data set

```{r}
distinct_set <- mtcars |> distinct(gear, carb, am)
distinct_set
```

sort it by increasing am, decreasing gear, decreasing carb d

```{r}
sorted_distinct_set <- distinct_set |> arrange(am, desc(gear), desc(carb))
sorted_distinct_set
```

Do this in two steps, get distinct working first, then arrange() it

# Column based Manipulation

#mutate

Alters columns, by adding new ones

Here is the power to weight (ptw) ratio

Note that here the altered version of mtcars is stored back into mtcars, so the
change in mtcars is kep

```{R}
#I accidentally ran this block and the next multiple times, so I was getting an error. Reloading mtcars fixes it.
data(mtcars)
mtcars=mtcars |> mutate(ptw=hp/wt)

```

#rename

renames a column

```{R}
mtcars=mtcars |> rename(power2weight=ptw)

head(mtcars)
```


# relocate

moves the location of a column

the default is to move variables to the front
.before and .after can alter the location relative to other named columsn

```{R}
mtcars=mtcars |> relocate(power2weight, .after=hp)
head(mtcars)
```

#*Question/Action*

Create a new column called specific_output

specific_output is hp divided by displacement

Place this behind power2weight in the table

Use head() to show this has worked.

```{r}
mtcars=mtcars |> mutate(specific_output=hp/disp)
mtcars=mtcars |> relocate(specific_output, .after=power2weight)
head(mtcars)
```

# Select

Just choices a set of columns

```{R}
mtcars |> select(cyl,carb,gear)
```
#*Question/Action*

Select specific_output, power2weight and qsec

```{r}
mtcars |> select(specific_output, power2weight, qsec)
```

#Groups and summarize

Summarize summarizes a specific statistic

We often want to group by one or more variables, then summarize

it looks like this

group_by() by cylinders and then compute mean mpg using summarize


```{R}
mtcars |>group_by(cyl) |> summarize(mean(hp))
```

we could then sort this by decreasing cylinders

```{R}
mtcars |>group_by(cyl) |> summarize(mean(hp)) |> arrange(desc(cyl))
```

We can group by multiple variables

and compute muliple outputs in the summary

```{R}
mtcars |> group_by(cyl,am) |>summarize (meanhp=mean(hp), meanpt2=mean(power2weight))|> arrange(desc(cyl))
```
This ability to combine group_by and summarize allows us to generate all kinds
of useful tables

Group_by cyl and carbs,  show the mean weight and mean mpg

```{r}
mtcars |> group_by(cyl,carb) |> summarize (meanwt=mean(wt), meanmpg=mean(mpg))
```

#*Question/Action*

Group by carb and cyl and find the mean power2weight ratio for each grouping

```{r}
mtcars |> group_by(carb, cyl) |> summarize (mean_power2weight=mean(power2weight))
```

#Counts

n=n() will give us the counts in the summary

```{R}
mtcars |> group_by(cyl,am) |>summarize (n=n())
```
we can also ask for the count of a variable to find n as well


```{R}
mtcars |> group_by(cyl,am) |>summarize(n=length(cyl)) |> arrange(desc(cyl))
```

# *Question/Action*

Make a table that shows both counts and mean horse power,   arranging by carb and gear
sort in descending carb and gear order

```{r}
mtcars |> group_by(carb,gear) |> summarize(n=n(), mean_hp = mean(hp)) |> arrange(desc(carb), gear)
```

# finding specific rows

df |> slice_head(n = 1) takes the first row from each group.
df |> slice_tail(n = 1) takes the last row in each group.
df |> slice_min(x, n = 1) takes the row with the smallest value of column x.
df |> slice_max(x, n = 1) takes the row with the largest value of column x.
df |> slice_sample(n = 1) takes one random row.

Find the max hp

```{R}
mtcars |> slice_max(hp,n=1)
```

#*Question/Action*

Find the min power2weight values for groupings by cylinder

```{r}
mtcars |> group_by(cyl) |> slice_min(power2weight,n=1)
```






