---
title: "Pair Programming Module 06 Classic Hypothesis Tests"
author: "HDS"
date: "2024-08-30"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Pair Programming Exercise 6  DSE5001

## Hypothesis tests based on Algebraic Tests

These are "classic" tests of various hypotheses.  We can't cover them all in 
depth this semester, but we can survey them.  These ideas crop up repeatedly 
throughout data science, so we need to be aware of them.

Most of these tests are about comparisons between groups.

We will use the mtcars data set

HD Sheets, August 30, 2024
checked 01/03/2025

# Student Info

Your name:Ryan Waterman
Your team-mates names: Trinity Tobin

# Libraries

```{R}
library("ggplot2")
library("dplyr")
```

# The mtcars data set

This is a nice data set as there are a variety of types of cars in it.

This is a set set from Motor Trend (MT) magazine from sometime in the mid 1970s,
so it is more or less historical data

```{r}
head(mtcars)
```

# Cyl as a factor

We want to think of cyl as factor, a group membership, so we will force
it to be a factor

When calling functions,  R will always treat variable as indicating group
membership if they are of type factor, so it is worth converting all group 
memberships as factors

```{R}
mtcars$cyl=factor(mtcars$cyl)

```

We will break the cars up into three categories based on the number of 
cylinders, 4,6 or 8

Use dplyrs to do this,  create separate data frames for 4,6,8 cylinder cars

```{R}
mtcars4<-mtcars %>% filter(cyl ==4)
mtcars6<-mtcars %>% filter(cyl==6)
mtcars8<-mtcars %>% filter(cyl==8)

```


## t-test for the Difference in the means of two groups

Suppose we state the hypothesis that 6 cylinder cars on average have more power than 4 cylinder cars.   This seems a fairly common sense idea

First, lets plot this, using a boxplot with color by cylinder using the whole data set

```{R}
ggplot(mtcars,aes(y=hp,x=cyl,color=cyl))+geom_boxplot()

```

Well, the plot looks pretty convincing, there is little or now overlap in the 
horsepower of 4 and 6 cylinder engines

# Running a formal t-test

```{R}
# we are sending in lists of hp values for 4 and 6 cylinders
# we specify that the alternative hypothesis is "greater", meaning the first
# variable value is hypothesized to be greater than the second

t.test(x=mtcars6$hp,y=mtcars4$hp,alternative="greater")
```
The results show us the mean values of x (6 cylinder hp) and of y (4 cylinder hp)

the observed t value was 3.5617,  or about 3.5 times the pooled standard deviation

with a degree of freedom of 11.49,  the value is 0.002088,  or about a 0.2% chance
the observed difference is due to chance

# *Question/Action*

Test the hypothesis that the mean weight of 8 cylinder cars is greater than the
mean weight of 6 cylinder cars, using a t-test

```{r}
t.test(x=mtcars8$wt,y=mtcars6$wt,alternative="greater")
```

*It appears that there is a 0.09% chance that the difference in mean weight by cylinder count is due to chance.*

# ANOVA,  differences among the means of more than two groups

Let's look at the mean hp of the 3 cyl categories

```{R}
mtcars %>% group_by(cyl) %>% summarize(mean_hp=mean(hp))

```

Looks like there are meaningful difference among all three
We can test this hypothesis using an ANOVA test

We are specifying a formula here, that hp is predicted by cyl

  this is written as   hp~cyl

while this called an ANOVA,  it is actually a form of generalized linear model (GLM)

```{R}
results_anova= aov(hp~cyl,data=mtcars)
summary(results_anova)
```

This test is giving us an F-statistic

The F is (variance explained by grouping by cylinder)/(average error variance)

F values are always ratios of variances,   high F means the grouping explains
a lot of differences in the data relative to the random fluctuation in the 
data

The p-value of an F ratio depends on the degrees of Freedom (DF),  the DF for
a factor is one less than the number of groups,   so we have 3 groups and the 
DF is 2

Typically F values over about 2 are meaningful, at reasonable sample sizes.

Here the p-value is very low, indicating no meaningful chance the variation in
hp is unrelated to the number of cylinders

# *Question/Action*

Determine if the differences in mean weights among 4,6 and 8 cylinder carries
is related to the cylinder category,  using an ANOVA as above

```{r}
results_anova= aov(wt~cyl,data=mtcars)
summary(results_anova)
```


## Tests of Counts in Categories

Contingency tables and Contingency tests (Chi-square)

The mtcar data set has a category for automatic or manual transmission.  the 
variable am = 0 for automatics and am=1 for manual

We can force am to be a factor and also set the labels names as "auto" and "manual"

```{R}
mtcars$am=factor(mtcars$am,labels=c("auto","manual"))
head(mtcars)
```
# Contingency table

Let's look at the counts of transmission type per cylinder, this is 
contingency table

```{R}

mtcars %>% group_by(cyl,am) %>% summarize(n=n())

```

We could also use the R function table to show the counts in each category

This is sometimes called a contingency table


```{R}
con_table=table(mtcars$cyl,mtcars$am)
con_table
```
We can get the sums over rows and columns

```{R}
library("MASS")

addmargins(con_table)
```
The Null hypothesis we might state is that the number of manual vs automatic transmissions
is independent of the number of cylinders.  As we can see in the table, that doesn't
look to be true, but the predominance of manual transmissions in 4 cylinder cares might just
be chance

The *Chi-squared test* is a test against the hypothesis of equal ratios in each cell
of the contingency table,  

Notice that we have to feed data into the test as a contingency table

```{R}
chisq.test(con_table)
```
The Chi-squared value is the sum of the squared departures from the expected count
in each cell, if the cells are randomly distributed among auto and manual

A large Chi-square is unlikely, the p value is determined by the df in the system

The df is (number of rows-1)*(number of columns-1)

This p value is low, so the chance of transmission type not being related to 
the number of cylinders is very low


The *Fisher exact test* is an alternative to the Chi-Square

We got a warning that the Chi-Square might not be a correct approximation,  the 
Chi-Square can fail at small sample sizes.

The Fisher exact test is more reliable,but hard to calculate at large sample 
sizes, so it is best suited to small sample sizes where the Chi-square is likely
to fail

```{r}
fisher.test(con_table)

```

The Fisher exact test returned a lower p-value than the Chi-squared, but the 
implication is the same.

## *Question/Action*

Create a contingency table for cylinders and gears.  

You will need to convert gears to a factor,   you don't need labels for the gears
though, the numbers are fine.

Show the table and then run the Chi-square and Fisher exact test.

Explain what it means.

```{r}
mtcar_factor<- mtcars %>% mutate(gear_factor = as.factor(gear))

con_table_cg=table(mtcar_factor$cyl,mtcar_factor$gear_factor)
con_table_cg

```

```{r}

chisq.test(con_table)
```

```{r}
fisher.test(con_table)

```

*The chi-squared test indicates a relationship between the number of factors and gears, as does the fisher test. Each result in a low p-value.*

# F test for equality of variance

In some tests, there is an assumption of equal variance among groups.  We may need to test this, or differences in variance may be our hypothesis

High variance in manufactured parts is a bad thing, so we may be looking to see if one set of parts had lower variance than another, as an example.

We can test to see if our three groups of cars by cylinder have equal variance in mpg

```{R}
mtcars %>% group_by(cyl) %>% summarise("Variance"=var(mpg))

```
Wow, sure looks like a no on that one!

In the 1970s, there were some really poor 4 cylinder engines

Here's the test

We can only compare two groups, so we'll compare 4 and 6, omitting 8

```{R}
mtcars46=mtcars[mtcars$cyl!=8,]

```

we have the formula mpg predicted by cyl, but we are testing variances, not means

In the var test we can only have two groups, so i used cyl==4, to compare
4 cylinders vs 6s an d8s

```{R}
var.test(mpg~cyl, data=mtcars46)
```
F is a ratio of variances, so high F means inequality

the p value depends on both df values,  here p=0.01182 for the null, so we
can reject the null

# *Question/Action*

Run an F-test for equality of variance on 6 cylinder vs 8 cylinder cars

```{R}
mtcars %>% group_by(cyl) %>% summarise("Variance"=var(mpg))

```

```{R}
mtcars68=mtcars[mtcars$cyl!=4,]

```

```{R}
var.test(mpg~cyl, data=mtcars68)
```
