---
title: "Pair_Programming_Long_data_Module3"
author: "HD Sheets"
date: "2024-08-12"
output: html_document
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

DSE5001 Module 3 pair programming exercise HD Sheets 8/13/2024 checked
01/03/2025

# Pair Programming Long Data, Module 3

We will look at the World Phones Data set

This data set needs a lot of work

-it doesn't start out as a data frame -the regions and years are labels,
not values -the rows and columns need to be flipped -we need to convert
it to long form

This is an example of "data wrangling" in which we need to do a lot of
data manipulation and structuring before we can do anything useful with
it.

Watch the steps needed to do this

-changing from one data storage form to another -changing data
types/formats -transposing-swapping rows and columns -more complex
changes

Load the data and look at it:

```{r}
data("WorldPhones")
WorldPhones

```

Okay, so what are the problems here?

This is not a particularly unusual table, but it' still a mess.

The variable being measured is the number of phones (in units of a
thousand phones)

They are recorded at different times and different locations

There is a composite key here, the region and the year with the measured
variable being the number of phones

What type of data storage is this? Use str() to find out what we are
dealing with

```{R}
str(WorldPhones)
```

I want to transpose this to flip the rows and columns and then put this
into a data frame

t()- transpose, converting rows t columns

data.frame()- convert from a numerical matrix to a dataframe

```{R}
phone_df=data.frame(t(WorldPhones))
phone_df

```

That did odd things to the column names, they have an X in them now

we can use rename to rename all the columns, sorta annoying but not hard

```{R}
library("tidyverse")

phone_df=phone_df |>rename("1951" = X1951,"1956"=X1956,"1957"=X1957,"1958"=X1958,"1959"=X1959,
                           "1960"=X1960,"1961"=X1961)

```

```{R}
phone_df
```

Right now, the regions are row labels, not variables. Dang.

Notice that in the list of regions, there is no listed column name, that
is because these values are not in a column, they are labels for each
row.

We need to add a column that is equal to the regions

I want to pivot longer and to do that the regions have to be in a
variable,

```{R}
phone_df=phone_df |> mutate(region=rownames(phone_df))
```

```{R}
phone_df
```

Now let's convert this to Long form

In the long form all the variables except region are being converted to
entries in the "year" column, with the associate values of those years
being stored in "phones"

This is an example of key-value storage. There is an identifier "region"
and then a key-value pair of the year (variable) and the number of
phones (the value)

```{R}
df_phones_long<-phone_df |> pivot_longer(!region,names_to="year",values_to="phones")

df_phones_long
```

Okay, that's much better

We can easily create some interesting visuals now

```{R}
library(ggplot2)

ggplot(df_phones_long,aes(x=year, y=phones,fill=region))+geom_bar(stat="identity",position="dodge")
```

```{R}
library(ggplot2)

ggplot(df_phones_long,aes(x=region, y=phones,fill=year))+geom_bar(stat="identity",position="dodge")
```

# *Question/Action*

The labels along the x-axis of the graph overlap each other and cannot
be read.

This is not "okay", you can't show anyone this graph like this.

We could fix the problem by figuring out how to rotate the labels along
the x-axis by ninety degrees.

ggplot allows for fine control of graph elements, such as the x-axis
label.

Google search and figure out how to rotate the x-axis labels by 90
degrees on this plot.\
Create a new code cell and enter the corrected R code to create the plot
above with the x-axis labels rotated by 90 degrees to make them readable.

```{r}
ggplot(df_phones_long,aes(x=region, y=phones,fill=year))+geom_bar(stat="identity",position="dodge") + theme(axis.text.x = element_text(angle = 90))
```


# *Question/Action*

Here is the Iris data set collected by Anderson and used in a famous
paper by RA Fisher

```{R}
data(iris)
head(iris)
```

# *Question/Action*

Do the following in a series of cells

-add the row number as data column, call it FlowerID

-convert this to long form, iris_long -you should have FlowerID and
species as your two keys -names_to should be "flower part" -values_to
should be "dimension"

-create a boxplot of values as y=dimension grouped by Species and
flowerpart This is a group boxplot using dimension and Species as the
grouping variables see
<https://r-graph-gallery.com/265-grouped-boxplot-with-ggplot2.html>

```         
       use x= Species and color=flowerpart
       
```

-reverse the grouping order above, so you have y=dimension grouped by
flowerpart and Species

```{R}
iris_df=iris |> mutate(FlowerID=rownames(iris))
```

```{R}
iris_df
```

```{R}
iris_df_long<-iris_df |> pivot_longer(!FlowerID & !Species,names_to="flowerpart",values_to="dimension")

iris_df_long
```

```{r}
ggplot(
  iris_df_long,
  mapping=aes(x=Species,y=dimension,fill=flowerpart)
       ) + geom_boxplot()
```

```{r}
ggplot(
  iris_df_long,
  mapping=aes(x=flowerpart,y=dimension,fill=Species)
       ) + geom_boxplot()
```

# Correlation

We'll work with the Iris data set again

We want just one species, not all of them, we'll just select setosa

We want to start with the wide data frame

```{R}
df_setosa= iris |> filter(Species=='setosa')

head(df_setosa)
```

Now let's look at correlation of Sepal.Length and Sepal.Width

Plotting first

```{R}
library("ggplot2")
ggplot(df_setosa,aes(x=Sepal.Length,y=Sepal.Width))+geom_point()

```

There looks to be a trend (ie correlation) here, with quite a bit of
noise

What is the correlation

```{R}
cor(df_setosa$Sepal.Length,df_setosa$Sepal.Width)

```

We have an R of 0.745, reasonably high but not extreme

We could look at the correlation of of Sepal length and width in all 3
species

```{R}
iris |>group_by(Species) |>summarize(R=cor(Sepal.Length,Sepal.Width))
```

#Looking at all Pairwise plots for the Setosa species data

The function ggpairs from GGally gives us a fast visual summary of the
data

We get histograms of each variable, boxplots of each variable, biplots
of each pair and the correlation of each pair

This is a handy tool for exploratory analysis, but is too much at once
for presentations


```{R}
library('GGally')
ggpairs(df_setosa)

```

# *Question/Action*

-Do all the distributions look normal/gaussian/bell curve? Explain why
or why not

*All but the petal width look normal. The petal width is deceiving because the resolution on the x-axis is very low, and one value had many instances, creating a false "bell curve".*

-Which biplots look like they show a trend?

*Sepal length vs. sepal width as well as petal length vs. petal width show some correlation.*

-Which two correlations are the highest?

*Sepal Length vs. Sepal Width and Petal Length vs Petal Width.*

-Which two variables seem to have the most outliers?

*Petal Length and Sepal Width. Sepal Width has a higher resolution along the x axis, so it likely has a higher quantity of outliers compared to petal width.*

# *Question/Action*

Load the mtcars built-in data set

```{r}
data(mtcars)
```


Select only mpg,disp, hp, wt and qsec from the data frame, call it
mtcars_few

```{r}
mtcars_few <- mtcars |> select(mpg, disp, hp, wt, qsec)
mtcars_few
```

Create a ggpairs plot

```{r}
ggpairs(mtcars_few)
```

-Which variables, if any, look normal?

*mpg, qsec, and it could be argued wt as well.*

-Which variables seem to have skew?

*All variables have skew.*

From the plots, which variables have positive correlation, which have
negatitve? Do any appear to have little or no correlation?

*Positive Correlation:*
*mpg vs. qsec, disp vs. hp, disp vs. wt, and hp vs. wt*

*Negative COrrelation:*
*mpg vs. disp, mpg vs. hp, mpg vs. wt, and hp vs. qsec*

*Little or No Correlation*
*qsec vs. wt*

-Which pair has the highest positive correlation?

*disp vs. wt*

-Which has the most extreme negative correlation?

*wt vs. mpg*

# *Question/Action*

Convert mtcars_few to a long version

```{R}
mtcars_few_long<-mtcars_few |> 
  mutate(Make_Model=rownames(mtcars_few)) |> 
  pivot_longer(!Make_Model,names_to="characteristic",values_to="value")

mtcars_few_long
```

Create a boxplot that shows all 5 variables in one plot

```{r}
ggplot(
  mtcars_few_long,
  mapping=aes(x=Make_Model,y=value,fill=characteristic)
       ) + geom_boxplot() + theme(axis.text.x = element_text(angle = 90))
```

*well... that certainly isn't ideal, let's select a smaller set of the data*

```{r}
mtcars_few_long<-head(mtcars_few_long,15)
```

```{r}
ggplot(
  mtcars_few_long,
  mapping=aes(x=Make_Model,y=value,fill=characteristic)
       ) + geom_boxplot() + theme(axis.text.x = element_text(angle = 90))
```

*I can't really argue that this is any better, there is just too much variance in the values across the characteristics.*



