---
title: "Pair_Programming_alternative_slicing"
author: "HDS"
date: "2024-08-13"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

HDS,  August 13, 2024
Checked 01/03/2025

# Alternative Slicing

The dplyr package has a lot of nice tools for sorting and slicing and summarizing
but there are other ways to do it.

In base R we can slice data frames using conditional statements, these
are often used to select both rows and columns

Last week we used Tidyverse tools to sort and slice,  we need some familiarity
with base R as well.

## Why discuss this

R almost always has at least three valid ways to do everything
The dplyr style slicing and sorting is really convenient, but dplyr is only
available in R,   in Python and other languages, slicing is done in the base R 
style.     

The base R style slicing was taken from linear algebra (aka matrix algebra), and 
appears in many other languages.

The tidyverse methods are derived from SQL-based methods of sorting and slicing.  These are
also seen in the Polars package for Python.

There is also a package called sqldf that allows you to use SQL queries on R dataframes.  If you
have a lot of experience with sql and want to leverage that,  look at sqldf:

  https://www.geeksforgeeks.org/how-to-write-a-sql-query-in-r/

We don't have time in DSE5001 to teach this, but if you do know SQL well, you can teach yourself sqldf pretty easily.


We will use mtcars as an example again to look at base R style manipulations.

```{R}
head(mtcars)
```

We can access portions of the dataframe using numeric ranges

We give the rows first, then the columns (this is a linear algebra convention
most languages follow,  rows, then columns)

```{R}
mtcars[1:3,1:5]
```

We can access columns by name

```{R}
mtcars[1:3,c("mpg","cyl","drat")]
```

We can access by a condition statement,  usually for rows, much like a filter
(well, identically to a filter)

```{R}
mtcars[mtcars$mpg>28,c("mpg","cyl")]
```
We can do more complex slices using logical OR (|) and  AND (&) operators


```{R}
mtcars[(mtcars$cyl==6)&(mtcars$hp>150),c("mpg","cyl")]
```

## Which

We can get the row numbers of variables that mean a specific condition
using the which operation

```{R}
which((mtcars$cyl==6)&(mtcars$hp>150))

```
we can then look at that row

leaving the columns blank gives us all columns

```{R}
mtcars[30,]

```

This is helpful looking for a minimum or maximum, for example

```{r}
max_hp_row=which(mtcars$hp==max(mtcars$hp))
mtcars[max_hp_row,]
```

# *Question/Action*

-Find the car with the highest mpq and print out it's row

```{r}
max_mpg_row=which(mtcars$mpg==max(mtcars$mpg))
max_mpg_row
```

-Find the cars with 8 cylinders and hp less than 150

```{R}
slice<-which((mtcars$cyl==8)&(mtcars$hp<150))
slice
mtcars[slice,]
```

*Trick question, there aren't any.*

## tapply

Tapply is an alternative to the summarize function in dplyr

It will apply any function to a column of a dataframe, grouped by one or more
other variables

we are applying the function mean() to the data mtcars$mpg, grouped by cyl and 
am


```{R}
tapply(mtcars$mpg,list(mtcars$cyl,mtcars$am),mean)
```

# *Question/Action*

Load the iris data 

```{r}
data(iris)
```

Find the mean value of Sepal.Length grouped by species

```{R}
tapply(iris$Sepal.Length,list(iris$Species),mean)
```

Find the standard deviation of Sepal.Length by species as well

```{R}
tapply(iris$Sepal.Length,list(iris$Species),sd)
```


